---
title: "Authentication and Project Setup"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Authentication and Project Setup}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

## Overview

`ukbflow` interacts with the UK Biobank Research Analysis Platform (RAP) via the DNAnexus `dx` toolkit. Authentication is required to access your UKB project, submit jobs, and retrieve results.

Two working modes are supported:

- **Local → RAP**: Run R locally, authenticate with a DNAnexus API token, and interact with RAP remotely.
- **RAP → RAP**: Run R inside the RAP cloud environment (e.g., RStudio on DNAnexus), where authentication is handled automatically by the platform.

---

## Obtaining an API Token

1. Log in to the [DNAnexus platform](https://platform.dnanexus.com)
2. Go to **Account Settings > API Tokens**
3. Click **New Token**, set an appropriate expiry, and copy the token

> Keep your token private. Treat it like a password — do not share it or commit it to version control.

---

## Storing Your Token Securely

The recommended approach is to store your token in `~/.Renviron`, a local file that is never committed to git. This keeps the token out of your scripts entirely.

```{r renviron}
usethis::edit_r_environ()
# Add the following line, then save and restart R:
# DX_API_TOKEN=your_token_here
```

After restarting R, the token is available to `auth_login()` automatically.

---

## Logging In

### Local → RAP

```{r auth-login}
library(ukbflow)

auth_login()  # reads DX_API_TOKEN from environment
```

The token is cached locally by the dx toolkit and persists across R sessions. **You only need to log in once** — subsequent R sessions do not require re-authentication unless the token expires or you explicitly log out.

If you prefer to pass the token directly (e.g., in an interactive session), you can do so:

```{r auth-login-direct}
auth_login("your_token_here")
```

> **Security note**: Avoid saving `auth_login("token")` calls in script files. Prefer the `.Renviron` approach for any workflow you intend to save or share.

### RAP → RAP

When running inside the RAP RStudio environment, authentication is handled automatically by the platform. No login step is required:

```{r auth-rap}
library(ukbflow)

auth_status()  # verify the current session
```

---

## Checking Authentication Status

```{r auth-status}
auth_status()
#> User:    Chris001127
#> Project: project-XXXXXXXXXXXX
```

`auth_status()` returns the current user and active project. Use this to confirm your session before running any analysis.

---

## Listing Available Projects

```{r list-projects}
auth_list_projects()
```

This returns all RAP projects accessible to your account, along with their project IDs. Project IDs take the form `project-XXXXXXXXXXXX` and are required for `auth_select_project()`.

---

## Selecting a Project

```{r select-project}
auth_select_project("project-XXXXXXXXXXXX")
#> Project selected: project-XXXXXXXXXXXX
```

`ukbflow` uses project IDs rather than names to avoid ambiguity. Use `auth_list_projects()` to find the correct ID first.

Once selected, the project context is saved by the dx toolkit and persists across sessions.

---

## Logging Out

```{r auth-logout}
auth_logout()
#> Logged out from DNAnexus.
```

Logging out clears the local session cache. The next call to any `auth_*` or `job_*` function will require re-authentication.

---

## Token Expiry

DNAnexus API tokens have a limited validity period set at creation time. When a token expires:

- `auth_login()` or other dx-dependent functions will fail
- Generate a new token from the DNAnexus platform and log in again:

```{r reauth}
auth_login("your_new_token_here")
```

Or update `~/.Renviron` with the new token and call `auth_login()` without arguments.

---

## Full Local → RAP Workflow

```{r full-workflow}
library(ukbflow)

auth_login()                                 # authenticate
auth_status()                                # verify session
auth_list_projects()                         # find your project ID
auth_select_project("project-XXXXXXXXXXXX")  # set active project

# ... run your analysis ...

auth_logout()                                # optional: clear session
```

---

## Getting Help

- `?auth_login`, `?auth_status`, `?auth_list_projects`, `?auth_select_project`, `?auth_logout`
- [DNAnexus documentation](https://documentation.dnanexus.com)
- [GitHub Issues](https://github.com/evanbio/ukbflow/issues)
