---
title: "Survival Analysis Setup for UKB Outcomes"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Survival Analysis Setup for UKB Outcomes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

## Overview

After disease case definitions have been derived (see `vignette("derive")`),
three additional functions prepare the data for time-to-event analysis:

| Function | Output columns | Purpose |
|---|---|---|
| `derive_timing()` | `{name}_timing` | Classify prevalent vs. incident disease |
| `derive_age()` | `age_at_{name}` | Age at event (years) |
| `derive_followup()` | `{name}_followup_end`, `{name}_followup_years` | Follow-up end date and duration |

> **Prerequisite**: `{name}_status`, `{name}_date`, `date_baseline`, and
> `age_at_recruitment` must already be present — produced by
> `vignette("derive")`.

---

## Step 1: Classify Timing — Prevalent vs. Incident

`derive_timing()` compares the disease date to the UKB baseline assessment
date and assigns each participant to one of four categories:

| Value | Meaning |
|---|---|
| `0` | No disease (`status` is `FALSE`) |
| `1` | **Prevalent** — disease date on or before baseline |
| `2` | **Incident** — disease date strictly after baseline |
| `NA` | Case with no recorded date; timing cannot be determined |

```{r derive-timing}
library(ukbflow)

# Uses {name}_status and {name}_date by default
df <- derive_timing(df, name = "outcome", baseline_col = "date_baseline")
#> ✔ derive_timing (outcome_timing):
#>   0 (no disease): 484 201
#>   1 (prevalent):   11 342
#>   2 (incident):     6 554
#>   NA (no date):         0
```

Supply explicit column names when the defaults do not apply:

```{r derive-timing-explicit}
df <- derive_timing(df,
  name         = "outcome_icd10",
  status_col   = "outcome_icd10",
  date_col     = "outcome_icd10_date",
  baseline_col = "date_baseline"
)
```

Call once per variable needed — for example, once for the combined case and
once per individual source (HES, self-report, etc.).

---

## Step 2: Age at Event

`derive_age()` computes age at the time of the event for cases, and returns
`NA` for non-cases and cases without a date.

$$\text{age\_at\_event} = \text{age\_at\_recruitment} + \frac{\text{event\_date} - \text{baseline\_date}}{365.25}$$

The divisor 365.25 accounts for leap years, ensuring sub-monthly precision in
age calculation across the full UKB follow-up window.

```{r derive-age}
# Multiple outcomes in one call — auto-detects {name}_date and {name}_status.
# Each name produces its own age_at_{name} column in a single pass.
df <- derive_age(df,
  name         = c("exposure", "outcome"),
  baseline_col = "date_baseline",
  age_col      = "age_at_recruitment"
)
#>   age_at_exposure: n=15 902, median=54.1, range=[40.2, 72.8]
#>   age_at_outcome: n=8 104, median=61.7, range=[40.8, 78.3]
#> ✔ derive_age: 2 events processed.
```

Supply explicit column mappings when names do not follow the default
`{name}_date` / `{name}_status` pattern:

```{r derive-age-explicit}
df <- derive_age(df,
  name         = "outcome_icd10",
  baseline_col = "date_baseline",
  age_col      = "age_at_recruitment",
  date_cols    = c(outcome_icd10 = "outcome_icd10_date"),
  status_cols  = c(outcome_icd10 = "outcome_icd10")
)
```

---

## Step 3: Follow-Up Time

`derive_followup()` computes the follow-up end date as the **earliest** of:

1. The outcome event date (if the participant is a case)
2. Date of death (field 40000; competing event)
3. Date lost to follow-up (field 191)
4. The administrative censoring date

Follow-up time in years is then derived from the baseline date.

```{r derive-followup}
df <- derive_followup(df,
  name         = "outcome",
  event_col    = "outcome_date",
  baseline_col = "date_baseline",
  censor_date  = as.Date("2022-06-01"),   # set to your study's cut-off date
  death_col    = "date_death",
  lost_col     = "date_lost_followup"
)
#> ✔ derive_followup (outcome):
#>   outcome_followup_end: 502 097 / 502 097 non-missing
#>   outcome_followup_years: mean=11.84, median=12.31, range=[0.03, 14.92]
```

Output columns:

| Column | Type | Description |
|---|---|---|
| `outcome_followup_end` | IDate | Earliest competing date |
| `outcome_followup_years` | numeric | Years from baseline to end |

### Prevalent cases receive `NA` follow-up time

Participants whose event date falls **before or on the baseline date** (prevalent
cases, `{name}_timing == 1`) will have `followup_years` set to `NA` rather than
a zero or negative value, which has no meaning in time-to-event analysis.
Use `derive_timing()` to identify and exclude prevalent cases before fitting a
Cox model (see the full pipeline example below).

### Auto-detection of death and lost-to-follow-up columns

When `death_col` and `lost_col` are `NULL` (default), `derive_followup()`
looks them up automatically from the field cache (UKB fields 40000 and 191).
Pass `FALSE` to explicitly disable a competing event:

```{r derive-followup-nodeath}
df <- derive_followup(df,
  name         = "outcome",
  event_col    = "outcome_date",
  baseline_col = "date_baseline",
  censor_date  = as.Date("2022-06-01"),
  death_col    = FALSE,
  lost_col     = FALSE
)
```

---

## Full Survival-Ready Pipeline

After completing all three steps, the data contains everything needed to fit
a Cox proportional hazards model:

```{r cox-example}
library(survival)

# Incident analysis: exclude prevalent cases and those with undetermined timing
df_incident <- df[outcome_timing != 1L]

fit <- coxph(
  Surv(outcome_followup_years, outcome_status) ~
    exposure_status + age_at_recruitment + sex + smoking_status_i0,
  data = df_incident
)
summary(fit)
```

Column roles in the model:

| Column | Role |
|---|---|
| `outcome_status` | Event indicator (logical) |
| `outcome_followup_years` | Time variable |
| `outcome_timing` | Filter: exclude prevalent (`== 1`) |
| `age_at_outcome` | Age at diagnosis (descriptive / secondary analysis) |
| `exposure_status` | Exposure of interest |

---

## Getting Help

- `?derive_timing`, `?derive_age`, `?derive_followup`
- `vignette("derive")` — disease phenotype derivation
- `vignette("decode")` — decoding column names and values
- [GitHub Issues](https://github.com/evanbio/ukbflow/issues)
