---
title: "Deriving Disease Phenotypes from UKB Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Deriving Disease Phenotypes from UKB Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

## Overview

The `derive_*` functions convert raw, decoded UKB columns into analysis-ready
variables. This vignette covers the disease phenotype derivation pipeline:

| Step | Function(s) | Purpose |
|---|---|---|
| 1 | `derive_missing()` | Handle "Do not know" / "Prefer not to answer" |
| 2 | `derive_covariate()` | Convert types; summarise covariates |
| 3 | `derive_cut()` | Bin continuous variables into groups |
| 4 | `derive_selfreport()` | Self-reported disease status + date |
| 5 | `derive_hes()` | HES inpatient ICD-10 status + date |
| 6 | `derive_first_occurrence()` | First Occurrence field status + date |
| 7 | `derive_cancer_registry()` | Cancer registry status + date |
| 8 | `derive_death_registry()` | Death registry ICD-10 status + date |
| 9 | `derive_icd10()` | Combine any subset of sources (wrapper) |
| 10 | `derive_case()` | Merge self-report + ICD-10 into final case definition |

All functions accept a `data.frame` or `data.table` and return a `data.table`
with new columns added **in-place** (no copy for `data.table` input).

> **Prerequisite**: data should have been extracted with `extract_pheno()` or
> `extract_batch()`, then decoded with `decode_values()` and `decode_names()`.
> See `vignette("decode")`.

---

## Extracting the Required Fields

The field IDs below are **universal across all UKB projects**. Include the
fields relevant to your analysis:

```{r extract}
library(ukbflow)

df <- extract_pheno(c(
  31,      # sex
  21022,   # age at recruitment
  53,      # date of baseline assessment (all visit instances)
  20116,   # smoking status
  # --- self-reported non-cancer illness ---
  20002,   # illness codes
  20008,   # year of diagnosis
  # --- self-reported cancer ---
  20001,   # cancer codes
  20006,   # year of cancer diagnosis
  # --- HES inpatient ---
  41270,   # ICD-10 code array
  41280,   # diagnosis date array
  # --- cancer registry ---
  40006,   # ICD-10 type
  40005,   # date of cancer diagnosis
  40011,   # histology
  40012,   # behaviour
  # --- death registry ---
  40001,   # primary cause of death
  40002,   # secondary cause of death
  40000,   # date of death
  # --- censoring / competing events ---
  191      # date lost to follow-up
)) |>
  decode_values() |>
  decode_names()

# Rename the verbose baseline date column to something concise
names(df)[names(df) == "date_of_attending_assessment_centre_i0"] <- "date_baseline"
```

---

## Step 1: Handle Informative Missing Labels

UKB uses special labels such as `"Do not know"` and `"Prefer not to answer"`
to distinguish refusal from true missing data. `derive_missing()` converts
these to `NA` (default) or retains them as `"Unknown"` for modelling.

```{r derive-missing}
df <- derive_missing(df)
#> ✔ derive_missing: replaced 4 219 values across 5 columns (action = "na").
```

> **Performance**: `derive_missing()` uses `data.table::set()` for in-place
> replacement — no column copies are made regardless of dataset size.
> Processing 500 000 rows typically completes in under a second with
> negligible memory overhead.

To keep non-response as a model category:

```{r derive-missing-unknown}
df <- derive_missing(df, action = "unknown")
```

To add custom labels beyond the built-in list:

```{r derive-missing-extra}
df <- derive_missing(df, extra_labels = "Not applicable")
```

---

## Step 2: Prepare Covariates

`derive_covariate()` converts character-encoded numeric columns to `numeric`
and categorical columns to `factor`, printing a distribution summary for each.

```{r derive-covariate}
df <- derive_covariate(df,
  as_numeric = "body_mass_index_bmi_i0",
  as_factor  = c("sex", "smoking_status_i0"),
  factor_levels = list(
    smoking_status_i0 = c("Never", "Previous", "Current")
  )
)
#> ── Numeric ──────────────────────────────────────────────────────────────────
#>   body_mass_index_bmi_i0: mean=27.4, median=26.8, sd=4.9, Q1=24.1, Q3=30.1, NA=0.9% (n=4512)
#> ── Factor ───────────────────────────────────────────────────────────────────
#>   sex [2 levels]
#>     Female: n=273 029 (54.4%)
#>     Male:   n=229 068 (45.6%)
#>   smoking_status_i0 [3 levels]
#>     Never:    n=271 400 (54.1%)
#>     Previous: n=150 736 (30.0%)
#>     Current:  n=79 867 (15.9%)
```

---

## Step 3: Bin Continuous Variables

`derive_cut()` creates a new factor column by binning a continuous variable
into quantile-based or custom groups.

```{r derive-cut-quantile}
# Equal-frequency tertiles (automatic breakpoints)
df <- derive_cut(df, col = "age_at_recruitment", n = 3)
# → adds age_at_recruitment_tri with levels Q1 / Q2 / Q3
```

```{r derive-cut-custom}
# Custom breakpoints with meaningful labels
df <- derive_cut(df,
  col    = "age_at_recruitment",
  n      = 3,
  breaks = c(50, 60),
  labels = c("<50", "50–59", "≥60"),
  name   = "age_group"
)
```

---

## Step 4: Self-Reported Disease

`derive_selfreport()` searches UKB self-reported non-cancer illness (field
20002) or cancer (field 20001) columns for a disease label matching a regex,
then returns binary status and the earliest report date.

```{r derive-selfreport}
# Non-cancer illness example
df <- derive_selfreport(df,
  name  = "disease",
  regex = "your disease name|synonym"
)
#> ✔ derive_selfreport (disease): 8104 cases, 8104 with date.

# Cancer illness example (field = "cancer")
df <- derive_selfreport(df,
  name  = "cancer_outcome",
  regex = "your cancer name",
  field = "cancer"
)
```

This adds two columns:

| Column | Type | Description |
|---|---|---|
| `disease_selfreport` | logical | `TRUE` if any instance matched |
| `disease_selfreport_date` | IDate | Earliest report date |

---

## Step 5: HES Inpatient Records

`derive_hes()` scans UKB Hospital Episode Statistics ICD-10 codes (field
41270, stored as a JSON array per participant) and matches the earliest
corresponding date from field 41280.

```{r derive-hes}
# Prefix match: codes starting with "E11" (e.g. E110, E119)
df <- derive_hes(df, name = "disease", icd10 = "E11")

# Exact match: only "E11" itself
df <- derive_hes(df, name = "disease", icd10 = "E11", match = "exact")

# Regex: full control over the pattern
df <- derive_hes(df, name = "disease", icd10 = "^E1[01]", match = "regex")
```

The `match` argument controls how codes are compared:

| `match` | Behaviour | Example |
|---|---|---|
| `"prefix"` (default) | Code starts with pattern | `"E11"` matches `"E110"`, `"E119"` |
| `"exact"` | Full 3- or 4-digit match | `"E11"` matches only `"E11"` |
| `"regex"` | Full regular expression | `"^E1[01]"` |

`match = "regex"` accepts full Perl-compatible regular expressions, making it
straightforward to capture related ICD-10 codes across chapters or subcategories
in a single call (e.g. `"^E1[01]"` captures both E10 and E11 simultaneously).

---

## Step 6: First Occurrence Fields

UKB First Occurrence fields (p131xxx) record the earliest date a condition
was observed across **all linked sources** — self-report, HES inpatient, GP
records, and death registry — pre-integrated by UKB. If your disease has a
corresponding 131xxx field, this is usually the fastest and most comprehensive
single-source case definition. Look up your disease in the
[UKB Field Finder](https://biobank.ndph.ox.ac.uk/showcase/search.cgi).

```{r derive-fo}
# Look up your disease's First Occurrence field ID in the UKB Field Finder,
# then replace fo_field_id accordingly.
fo_field_id <- 131000L   # <-- replace with your field ID

df <- derive_first_occurrence(df, name = "disease", field = fo_field_id)
#> ✔ derive_first_occurrence (disease): 12043 cases with valid date.
```

---

## Step 7: Cancer Registry

`derive_cancer_registry()` searches the cancer registry ICD-10 field (40006)
and optionally filters by histology (field 40011) and behaviour (field 40012).

```{r derive-cancer}
# ICD-10 only — no histology/behaviour filter
df <- derive_cancer_registry(df,
  name  = "cancer_outcome",
  icd10 = "^C50"   # breast cancer example
)
#> ✔ derive_cancer_registry (cancer_outcome): 2 318 cases.

# With histology and behaviour filters
df <- derive_cancer_registry(df,
  name      = "cancer_outcome",
  icd10     = "^C44",
  histology = c(8070, 8071, 8072),   # specify relevant morphology codes
  behaviour = 3L                      # 3 = malignant
)
```

---

## Step 8: Death Registry

`derive_death_registry()` searches primary (field 40001) and secondary (field
40002) causes of death for ICD-10 codes.

```{r derive-death}
df <- derive_death_registry(df, name = "disease", icd10 = "E11")
#> ✔ derive_death_registry (disease): 312 cases, 312 with date.
```

---

## Step 9: Combine Sources with `derive_icd10()`

`derive_icd10()` is a high-level wrapper that calls any combination of the
source-specific functions above and merges their outputs into a single status
column and earliest date.

```{r derive-icd10}
# Non-cancer disease: HES + death + First Occurrence
df <- derive_icd10(df,
  name     = "disease",
  icd10    = "E11",
  source   = c("hes", "death", "first_occurrence"),
  fo_field = fo_field_id   # defined above
)
#> ✔ derive_icd10 (disease): 14 287 cases across 3 sources, 14 287 with date.

# Cancer outcome: HES + cancer registry + death
df <- derive_icd10(df,
  name   = "cancer_outcome",
  icd10  = "^C50",
  match  = "regex",
  source = c("hes", "death", "cancer_registry")
)
```

Intermediate source columns are retained alongside the combined result:

| Column | Type | Description |
|---|---|---|
| `disease_icd10` | logical | `TRUE` if positive in any specified source |
| `disease_icd10_date` | IDate | Earliest date across all sources |
| `disease_hes` | logical | HES status |
| `disease_hes_date` | IDate | HES date |
| `disease_fo` | logical | First Occurrence status |
| `disease_fo_date` | IDate | First Occurrence date |
| `disease_death` | logical | Death registry status |
| `disease_death_date` | IDate | Death registry date |

---

## Step 10: Final Case Definition

`derive_case()` merges the self-report and ICD-10 flags into a unified case
status, with the earliest date across both sources taken via `pmin()`.

```{r derive-case}
df <- derive_case(df, name = "disease")
#> ✔ derive_case (disease): 15 902 cases, 15 902 with date.
#>   Both sources (disease_icd10 & disease_selfreport): 6 218
```

Output columns:

| Column | Type | Description |
|---|---|---|
| `disease_status` | logical | `TRUE` if positive in self-report OR ICD-10 |
| `disease_date` | IDate | **Earliest** date across all sources (`pmin`) |

> **Why the earliest date matters**: `disease_date` captures the true first
> occurrence of disease regardless of which source recorded it first. This
> column is the direct input to `derive_timing()`, `derive_age()`, and
> `derive_followup()` — it is the chronological anchor of every downstream
> survival analysis.

---

## Getting Help

- `?derive_missing`, `?derive_covariate`, `?derive_cut`
- `?derive_selfreport`, `?derive_hes`, `?derive_first_occurrence`
- `?derive_cancer_registry`, `?derive_death_registry`
- `?derive_icd10`, `?derive_case`
- `vignette("derive-survival")` — timing, age at event, follow-up
- `vignette("decode")` — decoding column names and values
- [GitHub Issues](https://github.com/evanbio/ukbflow/issues)
