---
title: "Exploring and Fetching RAP Files"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Exploring and Fetching RAP Files}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

## Overview

The `fetch_*` functions provide a convenient R interface for exploring and downloading files from your UK Biobank RAP project. Rather than switching to the terminal and using `dx` commands directly, you can browse your remote project structure and retrieve files entirely within your R session.

> **UK Biobank Data Policy (2024+)**: Only summary-level outputs and metadata files may be downloaded locally. Individual-level phenotype and genotype data must remain within the RAP environment.

---

## Prerequisites

Ensure you are authenticated before using any `fetch_*` functions:

```{r auth}
library(ukbflow)

auth_login()
auth_select_project("project-XXXXXXXXXXXX")
```

See `vignette("auth")` for details.

---

## Exploring Remote Files

### List files and folders

`fetch_ls()` lists the contents of a remote RAP directory, returning a structured data frame:

```{r fetch-ls}
# List project root
fetch_ls()
#>                  name   type    size            modified
#> 1  Showcase metadata folder    <NA>                <NA>
#> 2             results folder    <NA>                <NA>
#> 3        analysis.log   file  4.2 KB 2024-11-01 10:22:03

# List a specific folder
fetch_ls("Showcase metadata/")
#>          name  type     size            modified
#> 1   field.tsv  file  12.3 MB 2024-10-15 08:01:44
#> 2 esimpint.tsv  file   3.1 MB 2024-10-15 08:01:50

# Filter by type
fetch_ls("results/", type = "file")

# Filter by name pattern
fetch_ls("results/", pattern = "\\.csv$")
```

The returned data frame has four columns:

| Column | Description |
|---|---|
| `name` | File or folder name |
| `type` | `"file"` or `"folder"` |
| `size` | File size (e.g. `"1.2 MB"`), `NA` for folders |
| `modified` | Last modified time (`POSIXct`), `NA` for folders |

### Browse the directory tree

`fetch_tree()` prints a tree-like view of the remote project structure:

```{r fetch-tree}
# Top-level overview
fetch_tree()

# Drill into a subfolder
fetch_tree("results/", max_depth = 3)
```

> **Note**: Each level of recursion triggers one API call per folder. Keep `max_depth` at 2–3 for interactive use to avoid long waits on large projects.

---

## Generating Download URLs

`fetch_url()` generates temporary pre-authenticated HTTPS URLs for remote files. Useful for passing to downstream tools or scripting metadata and results workflows without triggering a full download.

```{r fetch-url}
# Single file
fetch_url("Showcase metadata/field.tsv")

# Entire folder (returns a named character vector)
fetch_url("Showcase metadata/", duration = "7d")
```

URLs are valid for the specified `duration` (default: `"1d"`).

---

## Downloading Files

### Single file or folder

`fetch_file()` downloads a file or an entire folder to a local directory.

> **Note**: `fetch_file()`, `fetch_metadata()`, and `fetch_field()` can only be called from within the RAP environment. Calling them locally will produce an error, as individual-level UKB data must remain on the platform.

```{r fetch-file}
# Download a single file
fetch_file("Showcase metadata/field.tsv", dest_dir = "data/")

# Download an entire folder
fetch_file("Showcase metadata/", dest_dir = "data/metadata/")

# Resume an interrupted download
fetch_file("results/summary_stats.csv", dest_dir = "data/", resume = TRUE)
```

Folders are downloaded in parallel using `curl::multi_download()` for efficiency.

### Metadata shortcuts

Two convenience wrappers are provided for commonly used UKB files:

```{r fetch-meta}
# Download all Showcase metadata files (field.tsv, encoding.tsv, etc.)
fetch_metadata(dest_dir = "data/metadata/")

# Download the field dictionary only
fetch_field(dest_dir = "data/metadata/")
```

---

## Common Options

`fetch_metadata()` and `fetch_field()` are thin wrappers around `fetch_file()`, so all three share the same download-control arguments:

| Argument | Default | Description |
|---|---|---|
| `dest_dir` | `"."` / `"data/metadata/"` | Local destination directory (created if needed). Default is `"data/metadata/"` for `fetch_metadata()` and `fetch_field()`. |
| `overwrite` | `FALSE` | Overwrite existing local files |
| `resume` | `FALSE` | Resume an interrupted download |
| `verbose` | `TRUE` | Show download progress |

---

## Getting Help

- `?fetch_ls`, `?fetch_tree`, `?fetch_url`, `?fetch_file`, `?fetch_metadata`, `?fetch_field`
- [GitHub Issues](https://github.com/evanbio/ukbflow/issues)
