---
title: "Get Started with ukbflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get Started with ukbflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

## Welcome to `ukbflow`

**`ukbflow`** is an R package for UK Biobank analysis on the
[Research Analysis Platform (RAP)](https://ukbiobank.dnanexus.com).
It covers the full midstream-to-downstream pipeline — from phenotype derivation
and association analysis to publication-ready figures and genetic risk scoring —
entirely within the RAP cloud environment.

## Installation

```{r install}
pak::pkg_install("evanbio/ukbflow")
```

## A Quick Taste

### Authenticate and extract data

```{r auth}
library(ukbflow)

auth_login()
auth_select_project("project-XXXXXXXXXXXX")

df <- extract_pheno(c(31, 21022, 53, 20116)) |>
  decode_values() |>
  decode_names()
```

### Derive a disease phenotype

```{r derive}
df <- df |>
  derive_missing() |>
  derive_icd10(name = "outcome", icd10 = "E11",
               source = c("hes", "first_occurrence")) |>
  derive_followup(name = "outcome", event_col = "outcome_date",
                  baseline_col = "date_baseline",
                  censor_date  = as.Date("2022-06-01"))
```

### Run an association model

```{r assoc}
res <- assoc_coxph(
  data         = df,
  outcome_col  = "outcome_status",
  time_col     = "outcome_followup_years",
  exposure_col = "exposure_status",
  covariates   = c("age_at_recruitment", "sex", "smoking_status_i0")
)
```

### Plot the results

```{r plot}
# plot_forest() takes a reshaped data.frame — see vignette("plot") for details
plot_tableone(df, vars = c("age_at_recruitment", "sex", "bmi"),
              strata = "outcome_status", dest = "results/table1")
```

## Full Function Overview

| Module | Key functions | Vignette |
|---|---|---|
| Auth | `auth_login()`, `auth_select_project()` | `vignette("auth")` |
| Fetch | `fetch_ls()`, `fetch_file()`, `fetch_tree()` | `vignette("fetch")` |
| Decode | `decode_values()`, `decode_names()` | `vignette("decode")` |
| Derive | `derive_missing()`, `derive_icd10()`, `derive_case()` | `vignette("derive")` |
| Survival | `derive_timing()`, `derive_age()`, `derive_followup()` | `vignette("derive-survival")` |
| Assoc | `assoc_coxph()`, `assoc_logistic()`, `assoc_subgroup()` | `vignette("assoc")` |
| Plot | `plot_forest()`, `plot_tableone()` | `vignette("plot")` |
| GRS | `grs_check()`, `grs_score()`, `grs_validate()` | `vignette("grs")` |

## End-to-End Case Study

For a complete worked example using a simulated UK Biobank cohort — covering
data loading, phenotype derivation, cohort assembly, Cox regression, and
publication-ready visualisation — see:

`vignette("smoking_lung_cancer")` — **Smoking and Lung Cancer Risk: A Complete Analysis Workflow**

## Additional Resources

- [Documentation site](https://evanbio.github.io/ukbflow/)
- [GitHub](https://github.com/evanbio/ukbflow)
- View all functions: `?ukbflow` or `help(package = "ukbflow")`

> *"All models are wrong, but some are publishable."*
>
> — after George Box
