---
title: "Genetic Risk Score (GRS) Analysis on UKB RAP"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Genetic Risk Score (GRS) Analysis on UKB RAP}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

## Overview

The `grs_*` functions provide an end-to-end pipeline for computing and
validating polygenic / genetic risk scores within the UK Biobank Research
Analysis Platform (RAP). Because individual-level genotype data cannot be
downloaded locally, all computationally intensive steps are executed as
cloud jobs via the DNAnexus Swiss Army Knife app.

| Function | Where it runs | Purpose |
|---|---|---|
| `grs_check()` | Local or RAP | Validate and export a SNP weights file |
| `grs_bgen2pgen()` | Submits RAP jobs | Convert UKB imputed BGEN files to PGEN |
| `grs_score()` | Submits RAP jobs | Score genotypes across chromosomes with plink2 |
| `grs_standardize()` | Local or RAP | Z-score standardise GRS columns |
| `grs_zscore()` | Local or RAP | Alias for `grs_standardize()` |
| `grs_validate()` | Local or RAP | Assess predictive performance (OR/HR, AUC, C-index) |

**Typical pipeline:**

```
grs_check()  →  grs_bgen2pgen()  →  grs_score()  →  grs_standardize()  →  grs_validate()
```

> **Prerequisite**: you must be authenticated to UKB RAP and have a project
> selected. See `vignette("auth")`.

---

## Step 1: Validate the Weights File — `grs_check()`

Before uploading to RAP, validate your SNP weights file. The function reads
the file, checks required columns, flags problems, and writes a
plink2-compatible space-delimited output.

**Required columns:**

| Column | Description |
|---|---|
| `snp` | SNP identifier (expected `rs` + digits format) |
| `effect_allele` | Effect allele: one of A / T / C / G |
| `beta` | Effect size (log-OR or beta); must be numeric |

```{r grs-check-local}
library(ukbflow)

# Local: weights file on your machine
w <- grs_check("weights.csv", dest = "weights_clean.txt")
#> Read 'weights.csv': 312 rows, 5 columns.
#> ✔ No NA values.
#> ✔ No duplicate SNPs.
#> ✔ All SNP IDs match rs[0-9]+ format.
#> ✔ All effect alleles are A/T/C/G.
#> Beta summary:
#>   Range     : -0.412 to 0.589
#>   Mean |beta|: 0.183
#>   Positive  : 187 (59.9%)
#>   Negative  : 125 (40.1%)
#>   Zero      : 0
#> ✔ Weights file passed checks: 312 SNPs ready for UKB RAP.
#> ✔ Saved: 'weights_clean.txt'
```

```{r grs-check-rap}
# On RAP (JupyterLab / RStudio) — use /mnt/project/ paths directly
w <- grs_check(
  file = "/mnt/project/weights/weights.csv",
  dest = "/mnt/project/weights/weights_clean.txt"
)
```

The function returns the validated `data.table` invisibly, which can be
passed directly to downstream functions.

---

## Step 2: Convert BGEN to PGEN — `grs_bgen2pgen()`

UKB imputed genotype data is stored in BGEN format on RAP. `grs_bgen2pgen()`
submits one Swiss Army Knife job per chromosome to convert BGEN files to the
faster PGEN format with a MAF filter applied via plink2.

This function is called from your local machine or RAP JupyterLab — the heavy
lifting runs entirely in the cloud.

```{r grs-bgen2pgen-test}
# Test on the smallest chromosome first
ids <- grs_bgen2pgen(chr = 22, priority = "high")
#> Uploading 'grs_bgen2pgen_std.R' to RAP ...
#> ✔ Uploaded: '/grs_bgen2pgen_std.R'
#> Submitting 1 job(s) — mem2_ssd1_v2_x4 / priority: high
#> ✔ chr22 → 'job-Gxxxxxxxxxxxxxxxxxxxxxxxx'
#> ✔ 1/1 job(s) submitted. Monitor with job_ls().
```

```{r grs-bgen2pgen-all}
# Full run: small chromosomes on standard, large on upgraded instance
ids_small <- grs_bgen2pgen(chr = 15:22)
ids_large <- grs_bgen2pgen(chr = 1:14, instance = "large")

# Monitor progress
job_ls()
job_wait(c(ids_small, ids_large))
```

**Instance types:**

| Preset | DNAnexus instance | Cores | RAM | SSD | Recommended for |
|---|---|---|---|---|---|
| `"standard"` | `mem2_ssd1_v2_x4` | 4 | 12 GB | 200 GB | chr 15–22 |
| `"large"` | `mem2_ssd2_v2_x8` | 8 | 28 GB | 640 GB | chr 1–14 |

**Key arguments:**

| Argument | Default | Description |
|---|---|---|
| `chr` | `1:22` | Chromosomes to process |
| `dest` | `"/pgen/"` | RAP output path for PGEN files |
| `maf` | `0.01` | MAF filter passed to plink2 `--maf` |
| `instance` | `"standard"` | Instance type preset |
| `priority` | `"low"` | Job priority (`"low"` or `"high"`) |

> **Storage warning**: chromosomes 1–14 may exceed the 200 GB SSD on
> `"standard"` instances. Use `instance = "large"` for these.

---

## Step 3: Calculate GRS Scores — `grs_score()`

Once PGEN files are ready, `grs_score()` uploads your weights file(s) and
submits one plink2 scoring job per GRS. Each job scores all 22 chromosomes
and saves a single CSV to RAP.

```{r grs-score}
ids <- grs_score(
  file = c(
    grs_a = "weights/grs_a_weights.txt",
    grs_b = "weights/grs_b_weights.txt"
  ),
  dest     = "/grs/",
  priority = "high"
)
#> ── Uploading 2 weight file(s) to RAP ────────────────────────────────────────
#> Uploading 'grs_a_weights.txt' ...
#> ✔ Uploaded: '/grs_a_weights.txt'
#> Uploading 'grs_b_weights.txt' ...
#> ✔ Uploaded: '/grs_b_weights.txt'
#> Submitting 2 job(s) — mem2_ssd1_v2_x4 / priority: high
#> ✔ grs_a → 'job-Gxxxxxxxxxxxxxxxxxxxxxxxx'
#> ✔ grs_b → 'job-Gyyyyyyyyyyyyyyyyyyyyyyyy'
#> ✔ 2/2 job(s) submitted. Monitor with job_ls().

job_wait(ids)
```

When running from RAP JupyterLab with weights already at the project root,
the upload step is skipped automatically:

```{r grs-score-rap}
# On RAP: weights already at /mnt/project/grs_a_weights.txt
ids <- grs_score(
  file = c(grs_a = "/mnt/project/grs_a_weights.txt"),
  dest = "/grs/"
)
#> ℹ grs_a_weights.txt already at RAP root, skipping upload.
```

> **Important**: the `maf` argument must match the value used in
> `grs_bgen2pgen()` so that the correct PGEN files are located.

Output per job: `/grs/<GRS_name>_scores.csv` with columns `IID` and the GRS
score named `GRS_<name>`.

---

## Step 4: Standardise GRS Columns — `grs_standardize()`

After downloading the score CSVs from RAP (via `fetch_file()`) and merging
them into your analysis dataset, Z-score standardise each GRS column so that
effect estimates are interpretable as per-SD associations.

```{r grs-standardize}
# Auto-detect all columns containing "grs" (case-insensitive)
cohort <- grs_standardize(cohort)
#> Auto-detected 2 GRS column(s): 'GRS_a', 'GRS_b'
#> ✔ GRS_a → GRS_a_z  [mean=0.0000, sd=1.0000]
#> ✔ GRS_b → GRS_b_z  [mean=0.0000, sd=1.0000]
```

```{r grs-standardize-explicit}
# Or specify columns explicitly
cohort <- grs_standardize(cohort, grs_cols = c("GRS_a", "GRS_b"))
```

`grs_zscore()` is an alias and produces an identical result.

The original columns are preserved; `_z` columns are inserted immediately
after their source column.

---

## Step 5: Validate Predictive Performance — `grs_validate()`

`grs_validate()` runs four complementary validation analyses for each GRS:

| Analysis | What it measures |
|---|---|
| **Per SD** | OR (logistic) or HR (Cox) per 1-SD increase in GRS |
| **High vs Low** | OR / HR comparing top 20% vs bottom 20% |
| **Trend test** | P-trend across Q1–Q4 quartiles |
| **Discrimination** | AUC (logistic) or C-index (Cox) with 95% CI |

### Logistic (cross-sectional)

```{r grs-validate-logistic}
res <- grs_validate(
  data        = cohort,
  grs_cols    = c("GRS_a_z", "GRS_b_z"),
  outcome_col = "outcome"
)
#> ── Creating GRS groups ───────────────────────────────────────────────────────
#> ── Effect per SD (OR) ───────────────────────────────────────────────────────
#> ── High vs Low ──────────────────────────────────────────────────────────────
#> ── Trend test ───────────────────────────────────────────────────────────────
#> ── AUC ──────────────────────────────────────────────────────────────────────
#> ✔ Validation complete.

res$per_sd
res$discrimination
```

### Cox (survival)

```{r grs-validate-cox}
res <- grs_validate(
  data        = cohort,
  grs_cols    = c("GRS_a_z", "GRS_b_z"),
  outcome_col = "outcome",
  time_col    = "followup_years",
  covariates  = c("age", "sex", paste0("pc", 1:10))
)

res$per_sd          # HR per SD × model
res$high_vs_low     # HR: top 20% vs bottom 20%
res$trend           # p-trend across Q1–Q4
res$discrimination  # C-index with 95% CI
```

**Return value:** a named list with four `data.table` elements.

| Element | Columns (logistic) | Columns (Cox) |
|---|---|---|
| `per_sd` | `GRS`, `model`, `OR`, `CI_lower`, `CI_upper`, `p_value` | same with `HR` |
| `high_vs_low` | same as `per_sd` | same with `HR` |
| `trend` | `exposure`, `model`, `p_trend` | same |
| `discrimination` | `GRS`, `AUC`, `CI_lower`, `CI_upper` | `GRS`, `C_index`, `CI_lower`, `CI_upper` |

> AUC calculation requires the `pROC` package.
> Install with `install.packages("pROC")` if needed.

---

## Complete Pipeline Example

```{r grs-full-pipeline}
library(ukbflow)

# 1. Validate weights (local)
grs_check("weights.csv", dest = "weights_clean.txt")

# 2. Convert BGEN → PGEN on RAP (submit jobs)
ids_std  <- grs_bgen2pgen(chr = 15:22, maf = 0.01)
ids_lrg  <- grs_bgen2pgen(chr = 1:14,  maf = 0.01, instance = "large")
job_wait(c(ids_std, ids_lrg))

# 3. Score GRS on RAP (submit jobs)
score_ids <- grs_score(
  file = c(grs_a = "weights_clean.txt"),
  maf  = 0.01,   # must match grs_bgen2pgen()
  dest = "/grs/"
)
job_wait(score_ids)

# 4. Download score CSV from RAP
fetch_file("/grs/GRS_a_scores.csv", dest = "GRS_a_scores.csv")

# 5. Merge into cohort and standardise
scores  <- data.table::fread("GRS_a_scores.csv")
cohort  <- scores[cohort, on = "IID"]
cohort  <- grs_standardize(cohort, grs_cols = "GRS_a")

# 6. Validate
res <- grs_validate(
  data        = cohort,
  grs_cols    = "GRS_a_z",
  outcome_col = "outcome",
  time_col    = "followup_years",
  covariates  = c("age", "sex", paste0("pc", 1:10))
)

res$per_sd
res$discrimination
```

---

## Getting Help

- `?grs_check`, `?grs_bgen2pgen`, `?grs_score`
- `?grs_standardize`, `?grs_validate`
- `vignette("auth")` — RAP authentication and project selection
- `vignette("job")` — monitoring and managing RAP jobs
- `vignette("assoc")` — association analysis functions used inside `grs_validate()`
- [GitHub Issues](https://github.com/evanbio/ukbflow/issues)
